---
title: Summarizing data
author: ~
date: '2018-03-08'
slug: summarizing-data
categories: []
tags: []
---

## Income

The data set we'll use to get familiar with summarizing data is a subset of the 1994 Census database. I'd encourage you to make a new Rscript (top left blank page with a + on it) and start by copying the following line which will grab the data from the UCI Machine Learning Repository.

```{r}
income <- read.csv("https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data", header = FALSE, stringsAsFactors = TRUE, 
                   col.names = c("age", "workclass", "samp.weight", "edu", "edu.num", "marital", "occ", "relationship",
                                 "race", "sex", "cap.gain", "cap.loss", "hrwk", "native", "income"), na.strings = c(" ?"))
```

If you want to take a peek at the data you can click on it in the environment (top right) pane after running that line of code. In general, a useful function in R is `str`. It's short for structure and as you might imagine shows us quite a bit of useful information about an object in a relatively small amount of output.

```{r}
str(income)
```

Here we can see that the `income` object is a `data.frame`. This is the primary form that data sets will come in for R. In addition though, we can see that we're working with a data set that has 32561 observations and we have information about 15 variables. In addition, for each of those 15 variables we get the name of that variable after the dollar sign, the type of variable it is (categorical variables are `factor` types in R and `int` stands for integer). We also get the first few observations for each variable! In practice, unless you're working with a data set with a ton of variables, `str` is probably the best place to start when you first get a data set. 

Another (slightly more crude) approach is to use the `head` function. This will give you the first few observations of a data set,

```{r}
head(income)
```

